\textbf{Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering}

### 2022-08-31 submission/manuscript.Rmd
shorted abstract and running title

Running title: Reference-based OTU clustering for machine learning
classification

@@

Machine learning classification of disease based on the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
*de novo* approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, such an approach depends on the sequences in the
dataset and therefore OTU assignments can change if new data are added.
This lack of stability complicates classification because in order to
use the model to classify additional samples, the new sequences must be
reclustered with the old data and the model must be retrained with the
new OTU assignments. The new reference-based clustering algorithm,
OptiFit, addresses this issue by fitting new sequences into existing
OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
whether this method for fitting new sequence data into existing OTUs
will impact the performance of classification models. We used OptiFit to
cluster additional data into existing OTU clusters and then evaluated
model performance in classifying a dataset containing samples from
patients with and without colonic screen relevant neoplasia (SRN). We
compared the performance of this model to the standard procedure of *de
novo* clustering all the data together. We found that both approaches
performed equally well in classifying SRNs. Moving forward, when OTUs
are used in classification problems, OptiFit can streamline the process
of classifying new samples by avoiding the need to retrain models using
reclustered sequences.

>>>>

Running title: Reference-based OTU clustering for ML classification

@@

Machine learning classification of disease using the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
de novo approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, such an approach depends on the sequences in the
dataset and therefore OTU assignments can change if new data are added.
This lack of stability complicates classification because in order to
use the model to classify additional samples, the new sequences must be
reclustered with the old data and the model must be retrained with the
new OTU assignments. The new reference-based clustering algorithm,
OptiFit, addresses this issue by fitting new sequences into existing
OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
whether this method for fitting new sequence data into existing OTUs
will impact the performance of classification models. We used OptiFit to
cluster additional data into existing OTU clusters and then evaluated
model performance in classifying a dataset containing samples from
patients with and without colonic screen relevant neoplasia (SRN). We
compared the performance of this model to the standard procedure of de
novo clustering all the data together. We found that both approaches
performed equally well in classifying SRNs. OptiFit can streamline the
process of classifying new samples by avoiding the need to retrain
models using reclustered sequences.

### 2022-08-31 submission/manuscript.tex
shorted abstract and running title

Machine learning classification of disease based on the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
\emph{de novo} approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, such an approach depends on the sequences in the
dataset and therefore OTU assignments can change if new data are added.
This lack of stability complicates classification because in order to
use the model to classify additional samples, the new sequences must be
reclustered with the old data and the model must be retrained with the
new OTU assignments. The new reference-based clustering algorithm,
OptiFit, addresses this issue by fitting new sequences into existing
OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
whether this method for fitting new sequence data into existing OTUs
will impact the performance of classification models. We used OptiFit to
cluster additional data into existing OTU clusters and then evaluated
model performance in classifying a dataset containing samples from
patients with and without colonic screen relevant neoplasia (SRN). We
compared the performance of this model to the standard procedure of
\emph{de novo} clustering all the data together. We found that both
approaches performed equally well in classifying SRNs. Moving forward,
when OTUs are used in classification problems, OptiFit can streamline
the process of classifying new samples by avoiding the need to retrain
models using reclustered sequences.

>>>>

Machine learning classification of disease using the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
de novo approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, such an approach depends on the sequences in the
dataset and therefore OTU assignments can change if new data are added.
This lack of stability complicates classification because in order to
use the model to classify additional samples, the new sequences must be
reclustered with the old data and the model must be retrained with the
new OTU assignments. The new reference-based clustering algorithm,
OptiFit, addresses this issue by fitting new sequences into existing
OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
whether this method for fitting new sequence data into existing OTUs
will impact the performance of classification models. We used OptiFit to
cluster additional data into existing OTU clusters and then evaluated
model performance in classifying a dataset containing samples from
patients with and without colonic screen relevant neoplasia (SRN). We
compared the performance of this model to the standard procedure of de
novo clustering all the data together. We found that both approaches
performed equally well in classifying SRNs. OptiFit can streamline the
process of classifying new samples by avoiding the need to retrain
models using reclustered sequences.
